Notes and other text

### 2023-04-02 doc/manuscript/translocation.qmd
Minor edits throughout Intro and Results

---
bibliography: translocation.bib
csl: pnas.csl
format: pdf
editor: visual
---

@@

The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains [California, USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
Endangered Species Act.
Although this decline was initiated by the introduction of non-native trout into the once-fishless portion of the Sierra Nevada [@bradford1989; @knapp2000], the arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.

@@

In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"? Mark here:  This transition reads fine to me. -->.
We then reintroduce these resistant frogs to 24 sites and describe the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.

@@

To assess whether frog populations show an evolutionary response to Bd exposure, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
Our goals were to identify putative genes and regions of the genome under selection and assess whether these genes could be related to increased frog resistance to Bd.
Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.

To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive vs. persistent) and individual variants.
We identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01).
Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$).
In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) <!--# This is a different definition than the one given in the Methods: "4 * standard deviation". -->.
Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions consistent with possible resistance mechanisms.

Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI) and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
Importantly, each cluster contained at least one population from both the naive and recovering groups (@fig-selectionresults B: SI) <!--# Add text indicating why this is important for interpretation of the selection analyses -->.
Our first analysis to detect signatures of selection identified a liberal set of variants that included 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs.
The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI; SI Dataset 1).
A single gene in both groups did not have an associated annotation.
For the SNPs in the stringent set of outliers, frequency differences between the naive and recovering populations ranged from 41% to 86%.
Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) <!--# Move this caveat text to the Discussion? :  and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors. -->, but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.
No biological functions, molecular functions, or cellular processes (i.e., GO terms) were significantly overrepresented in the set of 29 annotated outlier genes.

Our second analysis, using a splined window approach, identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
We identified 58 outlier windows for F~ST~ and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.
<!--# SNPs from the liberal or strict set? Also "windows" and "regions" are used interchangeably in this paragraph. To avoid confusion, use "windows" throughout? -->.
A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (SI Dataset 2).
The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.  Readers not familiar with amphibian-Bd systems might miss the importance of this result --> Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.

Together, our genomics results show that recovering populations have strong signatures of selection on multiple genes and regions of the genome.
While the functions of all genes and regions showing signatures of selection were not clearly interpretable biologically, regions of the genome associated with resistant phenotypes in other amphibian systems such as the extended MHC Class I, showed strong evidence of selection in recovering MYL frog populations.

@@

<!--# Based on Fig. 2, there are numerous outlier windows. How were the specific windows mentioned above chosen? Seems important to mention to avoid the perception that mentioned patterns were chosen haphazardly or cherry-picked based on unknown criteria. -->

<!--# In reading through these results, I am not seeing a compelling theme that emerges from all of the different results, both within and between the SNP and splined window sections. The results show differences in various metrics between population types, but are these changes large or small? I lack the context to understand that. Similarly, is the magnitude of differences consistent with strong or weak selection? In addition to the above, this section needs a concluding summary of the key genetic results. -->

@@

Comparison of genomes of frogs from Bd-naive versus recovering populations (described above) indicated substantial differences in allele frequencies, consistent with strong selection following Bd exposure.
Here we describe the outcome of frog reintroductions in which animals for reintroduction were collected from recovering populations, including two of the five recovering populations in the frog genome study (@fig-selectionresults: population 1 and 4).
<!--# Check this --> Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites [@joseph2018].
We refer to these reintroductions as "translocations" [@seddon2014].
During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap SI).
Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`) collected from 1 of `r n_distinct(d1$donor)` donor populations (@fig-yosemap SI).
Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens SI).
This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-viability-earlylate SI) <!--# Mention of viability is confusing because this issue isn't dealt with until the third section. Perhaps make the categorization based on frog survival? -->, which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R. sierrae*.
Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].
<!--# Add median Bd load, proportion below 5.8 copies threshold? --> <!--# The Methods-Translocation section contains text that repeats similar results. Edit to reduce repetition. --> <!--# To allow comparison with other translocation studies, adding information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation. -->

@@

Results from frog translocations indicated that adult survival was often relatively high and that nearly all populations showed evidence of successful reproduction (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \>5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).

We then included demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into our model described above and used it to predict long-term population viability.
Our model predicted that, following a single introduction of 40 adult individuals into a population, three of six populations with the highest adult survival probabilities had 50-year extinction probabilities less than 0.5 with our most conservative estimate of $\omega = 0.15$ (@fig-viability B).
As $\omega$ increased from 0.15 to 0.3, 50 year extinction probability was less 0.5 for all six populations, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
Overall, our model demonstrates that given observed yearly adult survival probabilities of resistant frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
This is likely a conservative estimate because there is evidence that naturally recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
Finally, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013.
Thus, there is some trend in our improved ability to select appropriate habitat to promote long-term viability of translocated, resistant populations.

@@

Disease-induced declines are a central conservation challenge for wildlife, and identifying biologically grounded strategies for assisted recovery is a critical need.
One potential strategy requires leveraging disease-induced genetic changes in host populations and so-called evolutionary rescue to recover a species at the landscape scale.
While previously proposed (Mendelson), this strategy has never been successfully implemented in wildlife disease systems.
Here, we provide strong evidence that selection for resistance and evolutionary rescue has led to the recovery of populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
Leveraging individuals from these resistant, rescued populations, 15 years of translocation effort and associated population viability modeling demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how natural evolutionary rescue can be used to successfully recover a species in the presence of disease.

>>>>

---
bibliography: translocation.bib
csl: pnas.csl
format: docx
editor: visual
---

@@

The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains [California, USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
Endangered Species Act.
In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000], and the arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.

@@

In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"? Mark here:  This transition reads fine to me. -->.
We then describe the reintroduction of these resistant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.

@@

To determine whether frog populations show an evolutionary response to Bd exposure, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
Our goals were to identify putative genes and regions of the genome under selection and assess whether these genes could be related to increased frog resistance to Bd.
Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.

To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants.
For these single nucleotide polymorphisms (SNPs), we identified outlier variants using two Bonferroni-corrected significance levels: "liberal" = 0.05, "stringent" = 0.01.
Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$).
In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) <!--# This is a different definition than the one given in the Methods: "4 * standard deviation". -->.
Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions, molecular functions, or cellular processes (i.e., GO terms) consistent with possible resistance mechanisms.

<!--# I suggest we mention the PCA in the paragraph above so it doesn't come as a surprise here. -->Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI), and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
Importantly, each cluster contained at least one population from both the naive and recovering groups (@fig-selectionresults B: SI) <!--# Add text indicating why this is important for interpretation of the selection analyses -->.
Our first analysis to detect signatures of selection identified a liberal set of variants that included 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs.
The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI; SI Dataset 1).
A single gene in both groups did not have an associated annotation.
For the SNPs in the stringent set of outliers, frequency differences between the naive and recovering populations ranged from 41% to 86%.
Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) <!--# Move this caveat text to the Discussion? :  and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors. -->, but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.
No GO terms were significantly over-represented in the set of 29 annotated outlier genes.

Our second analysis, using a splined window approach, identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.<!--# "windows" and "regions" are used interchangeably in this paragraph. To avoid confusion, use "windows" throughout? --> A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2 <!--# In the supplementary_table_outliers_typofix.xlsx file > the splined_window_shared_ouliers sheet contains descriptions of 10 shared regions, not 9-->) and no GO terms were overrepresented within this set of genes.
The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$), indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.  Readers not familiar with amphibian-Bd systems might miss the importance of this result --><!--# Based on Fu and Waldman (2017), are any of our data sufficient to conclude anything related to statement that "Bd-resistant species share similar pocket conformations within the MHC-II antigen-binding groove. Among susceptible species, survivors of epizootics bear alleles encoding these conformations."? What about MHC Class II characteristics linked to resistance by Bataille et al. (2015)? Savage and Zamudio (2011, 2016), Kosch et al. (2018)?-->

Together, our genomics results show that recovering populations have strong signatures of selection on multiple genes and regions of the genome.
While the functions of all genes and regions showing signatures of selection were not clearly interpretable, regions of the genome associated with resistant phenotypes in other amphibian systems such as the MHC complex, showed strong evidence of selection in recovering MYL frog populations.

@@

<!--# Based on Fig. 2, there are numerous outlier windows. How were the specific windows mentioned in above and in Fig. 2 C, D chosen? Seems important to mention to avoid the perception that mentioned patterns were chosen haphazardly or cherry-picked based on unknown criteria. -->

<!--# In reading through these results, I am not seeing a compelling theme that emerges from all of the different results, both within and between the SNP and splined window sections. The results show differences in various metrics between population types, but are these changes large or small? I lack the context to understand that. Similarly, is the magnitude of differences consistent with strong or weak selection? -->

@@

Comparison of genomes of frogs from Bd-naive versus recovering populations (described above) indicated substantial differences in allele frequencies and genomic regions, consistent with strong selection following Bd exposure.
In this section, we describe the outcome of frog reintroductions in which animals for reintroduction were collected from recovering populations, including two of the five recovering populations in the frog evolution study (@fig-allelefrequencies A: population 1 and 4).
<!--# Check this --> Reintroductions of *R. sierrae* were conducted as part of an effort to reestablish populations of this endangered species.
These reintroductions were conducted in Yosemite National Park (California, USA) and involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites [@joseph2018].
We refer to these reintroductions as "translocations" [@seddon2014].
During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap SI).
Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`) collected from 1 of `r n_distinct(d1$donor)` donor populations (@fig-yosemap SI).
Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens SI).
This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-viability-earlylate SI) <!--# Mention of viability is confusing because this issue isn't dealt with until the third section. Perhaps make the categorization based on frog survival? -->, which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R. sierrae*.
Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].
<!--# Add median Bd load, proportion below 5.8 copies threshold? --> <!--# The Methods-Translocation section contains text that repeats similar results. Edit to reduce repetition. --> <!--# To allow comparison with other translocation studies, adding information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation. -->

@@

Results from frog translocations indicated that adult survival was often relatively high and that nearly all populations showed evidence of successful reproduction (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \> 5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).

We then included demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into our model described above and used it to predict long-term population viability.
<!--# In regard to the previous sentence, is it worth stating that the inclusion of demographic stocahsticity and environmental variability will not change estimates of lambda? That might not be apparent. --><!--# In the following sentences, I edited the text to present the results based on omega = 0.3 first and then the results with reduced omega of 0.15. This matches the order in which we present the results in the previous paragraph. -->Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, all six populations with the highest adult survival probabilities had 50-year extinction probabilities less than 0.5 (<!--# provide actual range here -->@fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
When we decreased $\omega$ from 0.3 to our most conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.5 to 0.X <!--# give actual range here -->for three additional populations (@fig-viability B).
In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
Finally, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013.
Thus, there is some trend in our improved ability to select appropriate habitat to promote long-term viability of translocated, resistant populations.

@@

Disease-induced declines are a central conservation challenge for wildlife, and identifying strategies to recover affected species is a critical need.
One potential strategy, and the one we use in the current study, leverages disease-induced genetic changes in host populations and evolutionary rescue to recover species at the landscape scale.
Although previously proposed [@mendelson2019], to our knowledge this strategy has never been successfully implemented in wildlife disease systems [but see @joseph2018 for a small-scale example].
Here, we provide strong evidence that natural selection for resistance and the associated evolutionary rescue has facilitated recovery of populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
The use of individuals from these resistant, rescued populations in a 15 year translocation effort plus associated population viability modeling demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how evolutionary rescue can be used to successfully recover a species in the presence of disease.

Effectively leveraging evolutionary rescue for landscape-level species recovery requires strong evidence that evolutionary rescue is in fact the mechanism underlying recovery.
In the MYL frog system, there are four lines of evidence that support evolutionary rescue as primary mechanism of recovery.
First, in host-pathogen systems evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain permanent immunity to Bd and are susceptible to re-infection (Knapp, PeerJ), and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) <!--# This first line of evidence is weak and could be dropped. Just because conditions under which ER is most likely exist in MYL frogs doesn't mean much regarding whether ER is the most likely mechanism underlying recovery. -->.
Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering populations.
These differences are consistent with strong selection and some regions of the genome under selection are associated with known cellular mechanisms associated with resistance, such those related amphibian immune responses <!--# Add specifics to make this stronger, broaden to include not just allele frequencies but genomic regions. -->.
Third, our translocations and long-term assessment of establishment dynamics demonstrate that evolved resistance extends beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frogs and recruits into the adult population in the presence of Bd.
Finally, resistance can be statistically quantified --- in recovering populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics) <!--# Whether this resistance is due to evolution or adaptive immunity is not distinguishable under this line of evidence. -->.
Although other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), few have demonstrated evolutionary rescue: disease-induced selection of host resistance and subsequent host recovery.
Our study is unique in that we provide strong evidence that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.
